---
title: "107.106 AKSTA Statistical Computing"
subtitle: "Data Objects"
author: "Fabian Schroeder"
date: "SS 2018"
#output: slidy_presentation

output: html_document
---

***

\Large

> \ *Everything that exists is an object.*    
> \ *Everything that happens is a function call.*  
> \ *(John Chambers)*

\normalsize  

***  


## Data Objects

<style>
slides > slide { overflow: scroll; }

</style>

In every computer language variables provide a means of accessing the data stored in memory. R does not provide direct access to the computer’s memory but rather provides a number of specialized data structures we will refer to as objects. These objects are referred to through symbols or variables.

There are many different objects. In this lesson we will focus on those kinds that are ment to store data: *vectors*, *matrices*, *arrays*, *lists*, and *data frames*. These data objects differ with respect to their dimensionality and whether they are homogenous or heterogenous. 

|  | homogenous | heterogenous |
| -- | ---- | ----- |
| 1D | vector | list |
| 2D | matrix | data frame|
| nD | array | |


For every object we will learn how initialize it, fill it with data, index it, and learn about its *attributes*.


## Vectors

Vectors are one-dimensional data structures. In this section we will cover *atomic vectors*, where all elements must be of the same *type*.

### Initialize and Create Vectors

```{r collapse = TRUE}
# initialize a vector
vec_num <- vector(mode = "numeric", length = 7)
vec_log <- vector(mode = "logical", length = 7)
vec_chr <- vector(mode = "character", length = 7)
```

In order to obtain the default values of the vector() function one can simply print them to the console
```{r collapse = TRUE}
vec_num
vec_log
vec_chr
```

The following functions generate vectors and fill them according to some rule.
```{r collapse = TRUE}
# concatenate c()
vec_chr <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
vec_int <- c(1L, 0L, 1L, 2L, 3L, 0L, 2L)
vec_dou <- c(1.0, 0, 1.0, 2.0, 3.0, 0, 2.0)
is.vector(vec_int)

# seqence seq()
vec_num <- seq(1,7)
is.vector(vec_num)

# repeat rep()
vec_log <- rep(TRUE, 7)
is.vector(vec_log)
```

Vectors have three different properties: a *type* (`typeof()`), a length (`length()`), and its attributes including the names. R has six basic (‘atomic’) vector types: logical, integer, double, string (or character).  

```{r collapse = TRUE}
typeof(vec_log)
typeof(vec_int)
typeof(vec_dou)

length(vec_num)

names(vec_num) <- vec_chr
attributes(vec_num)

# use the structure function
str(vec_num)
str(vec_int)

```

For programming also the following functions can be used to check the type of a vector: `is.logical()`, `is.character()`, `is.numeric()`, `is.double()`, `is.atomic()`, `is.numeric()` .



### Indexing

In R there is a wide range of indexing options. Even for atomic vectors there  are **six** ways to subset:

1. **positive indexing** using positive integers.
2. **negative indexing** using negative integers.
3. **logical indexing** using logical vectors.
4. **named indexing** using character vectors.
5. **blank** indexing.
6. **zero** indexing.

```{r collapse = TRUE}
# positive indexing
vec_num[2]

# positive indexing using a vector
vec_num[c(1,3)]

# negative indexing
vec_num[-c(2,4)]

# named indexing
vec_num["Wednesday"]

# blank indexing
vec_num[]

# zero indexing
vec_num[0]

```

Logical indexing is expecially useful because we can use logical operators to subset vectors.

```{r collapse = TRUE}
vec_num[c(FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, TRUE)]

vec_num[vec_num == 0]

vec_num[vec_num == 2]
```


## Matrices and Arrays

Matrices and Data Frames are generalisation of vectors in the sense that they have two or more dimensions. They are initialized by means of the `matrix()` and the `array()` functions. 

```{r collapse = TRUE}
mat <- matrix(1:6, ncol=3, nrow=2)
mat

arr <- array(1:24, dim=c(3,4,2))
arr

```

Matrices and Arrays have different attributes





Subsetting of higher dimensional objects can be done in three ways:

1. using multiple vectors.
2. using a single vector.
3. using matrices

Subsetting multi-dimensional structures generally follows the same rules as single-dimensional
indexing for each index variable, with the relevant component of dimnames taking the place of
names. A couple of special rules apply, though:
Normally, a structure is accessed using the number of indices corresponding to its dimension.
It is however also possible to use a single index in which case the dim and dimnames attributes
are disregarded and the result is effectively that of c(m)[i]. Notice that m[1] is usually very
different from m[1, ] or m[, 1].
It is possible to use a matrix of integers as an index. In this case, the number of columns
of the matrix should match the number of dimensions of the structure, and the result will be a
vector with length as the number of rows of the matrix. The following example shows how to
extract the elements m[1, 1] and m[2, 2] in one operation.

```{r collapse = TRUE}
mat[c(1,4,6)]

mat[1,3]

mat[1,]

mat[,2]

mat[matrix(rep(c(TRUE, FALSE), 3), nrow = 2)]

arr[c(2,12,24)]

arr[1,2,1]

arr[1,,]

arr[array(rep(c(TRUE, FALSE), 12), dim = c(3,4,2) )]

```

Matrices have a few important attributes.

```{r collapse = TRUE} 
ncol(mat)
nrow(mat)

colnames(mat) <- LETTERS[1:3]
rownames(mat) <- letters[1:2]
mat

# you can also use
dimnames(mat)

```

The counterpart of length for an array is dim and the counterpart
of names is dimnames which is list of character vectors of
appropriate length.

```{r collapse = TRUE}
dim(arr)
dimnames(arr) <- list( c("r1", "r2", "r3"), c("c1", "c2", "c3", "c4"),c("a1", "a2"))

arr
```

The extension for `c()` for matrices is `cbind()` and `rbind()`. Similarly, the package abind provides the function `abind()`. For transposing a matrix in R the function `t()` is available and for the array counterpart the function aperm. To check if an object is a matrix / array the functions `is.matrix()` / `is.array()` can be used. Similarly coercion to matrices and arrays can be performed using `as.matrix()` / `as.array()`.


## Lists

Lists (“generic vectors”) are another kind of "one-dimensional" data objects. However, while in an atomic vector all elements must be of the same *type*, a list can contain elements of different types. In fact, lists have elements, each of whichcan contain any type of R object, i.e. the elements of a list do not have to be of the same type. A list can even contain vectors or lists.

A list is initialized by the `list()` function.

```{r collapse = TRUE}
# initialize a list
lst <- list("Logical_Vector" = vec_log, 
             "Integer_Vector" = vec_int, 
             "Double_Vector" = vec_dou,
             "Character_Vector" = vec_chr, 
             "Result" = TRUE)


```

Use the structure funtion to get a detailed overvier over its elements

```{r collapse = TRUE}
str(lst)
```

List elements are accessed through three different indexing operations. 

1. `[ ]`
2. `[[ ]]`
3. `$`

```{r collapse=TRUE, error=TRUE}
lst[1]

lst[[1]]

lst$Logical_Vector
```

The first one returns subsets a list, the other two options extract list components. The difference is subtle. A subset of a list is always a list itself, possibly of length 1. 

```{r collapse = TRUE }
class(lst[1])

class(lst[[1]])

class(lst$Logical_Vector)

```

Just like vectors, lists can be concatenated using the `c()` function. 

Lists have the same attributes as atomical vectors: length, names, ty


## Factors

Factors are used to describe items that can have a finite number of values (gender, social class,
etc.). Factors are currently implemented using an integer array to specify the actual *levels* and
a second array of names (*labels*) that are mapped to the integers. Factors have a class attribute which has the value *factor* which makes it behave differently compared to standard integer vector.

A factor is usually created with the function factor.

```{r collapse = TRUE, error = TRUE}
fac <- factor(c("green", "green", "blue", "green", "blue"))
```

A factor has its own attributes

```{r collapse = TRUE, error = TRUE}
class(fac)
levels(fac)
labels(fac)
```

Factors are indexed in the same manner atomic vectors are indexed. However, factor vectors cannot be concatenated or assigned values like atomic vectors. 

```{r collapse=TRUE}
# assignment idiosyncracies
fac[2] <- "green"
fac[2] <- "red"

# concatenation idiosyncracies
fac_1 <- factor(c("green", "green", "blue"))
fac_2 <- factor(c("green", "blue", "blue"))

c(fac_1, fac_2)
factor(c(fac_1, fac_2), labels = c("green", "blue"))
```




## Data Frames

Data frames are the most widely used structures in statistics. A data frame is a list of vectors, factors, and/or matrices all having the same length (number of rows in the case of matrices). It is, thus, a "two-dimensional" and shares some characteristics of matrices. Hence it shares properties of both vectors and matrices.

A *data frame* is initialized by means of the function `data.frame()`. 

```{r collapse = TRUE}
df <- data.frame(V1 = 1:5, V2 = c("a", "a", "b", "a", "d"))
str(df)
```

Since data frames are a mixture between lists and matrices, they have the attributes of both. For example the function `length()` returns the number of list components, which is the number of columns and hence the same as `ncol()`. While `nrow()` returns the the number of rows. Following the same reasoning, `names()` gives the names of the vectors which is the same as colnames. `rownames()` in turn gives the rownames.

```{r collapse = TRUE}
length(df)
dim(df)

names(df)
colnames(df)
rownames(df)
```

Lists, vectors and matrices can be coerced to data frames if it is
appropriate. For lists this means if not all objects have the same
“length”.

```{r collapse = TRUE}
vec_1 <- 1:5
lst_1 <- list(vec_1 = vec_1, vec_2 = letters[c(1,2,3,2,1)])
lst_2 <- list(vec_1 = vec_1, vec_2 = letters[c(1,2,3,2,1,3)])
str(as.data.frame(vec_1))

str(as.data.frame(mat))

str(as.data.frame(lst_1))
```

The basic functions to combine two data frames (works similar with matrices) are cbind and rbind. When combining column-wise, then the numbers of rows must match and row names are ignored (hence observations need to be in the same order). When combining row-wise. The number of columns and their names must match. For more advanced combining see the function merge.


## NULL

There is a special object called NULL. It is used whenever there is a need to indicate or specify
that an object is absent. It should not be confused with a vector or list of zero length.
The NULL object has no type and no modifiable properties. There is only one NULL object
in R, to which all instances refer. To test for NULL use is.null. You cannot set attributes on
NULL.

```{r chunk02, echo = TRUE, eval = FALSE}

a <- NULL
typeof(a)
mode(a)
storage.mode(a)

```
