---
title: "107.106 AKSTA Statistical Computing"
subtitle: "Functions and Operators"
author: "Fabian Schroeder"
date: "SS 2018"
output: html_document
---
 
***

\Large

> \ *Everything that exists is an object.*    
> \ *Everything that happens is a function call.*  
> \ *(John Chambers)*

\normalsize  

***  

## Functions

Functions are **fundamental building blocks** in R since they can be used to automate common tasks. As a rule of thumb, one should use a function whenever a block of code is evaluated more than twice with different parameters. The use of function makes the code much more structered and readable and, thus, avoids mistakes. 

The syntax for defining a function in R is 

```{r}
func_name <- function(arg1, arg2, ...) {
  statements
  return(object)
}

```


> Excercise 1: Implement the Fibonacci series using a recursive function.

```{r }

fibo_rec <- function(n) {
  
  if (n <= 2) {
    return(1)
  } else {
    return(fibo_rec(n-1) + fibo_rec(n-2))
  }
  
}

```

### Structure of Functions

All R functions have three components:

1.  `body`: the statements inside the function.
2.  `formals`: the list of arguments which specify how to call the function.
3.  `environment`: a guide to where variables of the function are located. When printing the function in the console it will display these parts. The environment is not shown if the function live in the global environment.

The information about the body, the formals, and the environment of a function can be accessed by simply typing the name of the function into the console without the brackets.

```{r collapse = TRUE, error = TRUE}
sd
```

This also holds true for your own functions. We can use the functions `formals()`, `body()`, and   `environment()` to print this information to the console.

```{r collapse = TRUE, error = TRUE}
f <- function(x, y = 1) 2*(x+y)

formals(f)

body(f)

environment(f)
```

There is one exception of a group of functions which do not have the three parts just described - these are called **primitive functions**. All primitive functions are located in the **base** package and call directly **C code** and do not contain any R code. Thus, the `formals()`, the `body()`, and the `environment()` of primitive function are all `NULL`.

```{r collapse=TRUE}
sum
formals(sum)
body(sum)
environment(sum)
```


> Excercise B: Make a list of all function of the base package

```{r eval = FALSE}
objs <- mget(ls("package:base"), inherits = TRUE)
funs <- Filter(is.function, objs)
names(funs)
```

> Excercise C: Select all primitive functions from the base package

```{r eval = FALSE}
prim.funs <- Filter(is.primitive, funs)
names(prim.funs)
```


### Function arguments

One has to distinguish between two types of arguments: 

1. Formal arguments: these arguments are the "property" of the function.
2. Actual arguments: these arguments can change with the function call.

Hence, to use functions effectively, one needs to understand how the actual arguments are matched with the formal arguments. There are different ways to call functions:

1. Named argument call: Arguments are matched by exact names.
2. Partially named argument call: Arguments are matched using the shortest unique string.
3. Positioning argument call: using the position of the arguments in the function definition.

The three different ways can also be mixed in a function call. Then R uses first named matching, then partial named matching and finally position matching.

```{r eval=FALSE, collapse=TRUE, error=TRUE}
f <- function(Position1, Pos2, Pos3) 
  {
  list(pos1=Position1, pos2=Pos2, pos3=Pos3)
  }

# named argument call
str(f(Position1 = 1, Pos2 = 2, Pos3 =3))

# when using a named argument call, the order doesn't matter
str(f(Pos2 = 2, Position1 = 1, Pos3 =3))

# positioning argument call
str(f(1, 2, 3))

# mixing argument call styles
str(f(2, 3, Position1 = 1))

# partially named argument calls
str(f(Posi = 1, 2, 3))
```

```{r collapse=TRUE, error=TRUE}
str(f(1, Pos = 2, 3))
```

1. Positional matching should be only used for the very few first arguments. This is also why in 
R functions the most important arguments to be specified by the user are always at the beginning.
2. Partial matching should be avoided when possible when programming. For example packages submitted to CRAN cannot use partial matching.
3. For formal arguments appearing after the special argument `...`, which will be discussed later, only (exact) named matching is possible. This is one reason why `...` is usually the last argument.  

```{r collapse=TRUE, error=TRUE}
# good calls
mean(1:10)
mean(1:10, trim=0.2)

# confusing calls
mean(1:10, n=T)
mean(1:10, , FALSE)
mean(1:10, 0.2)
mean(, , TRUE, x=1:10)
```


In programming you will often have the case, that the arguments to call a function are specified in list.
How to use that then for the actual function call? The solution is the function `do.call()`.

```{r collapse=TRUE, error=TRUE}
L1 <- list(1:10, na.rm=TRUE)
do.call(mean, L1)
```

Most functions in R give **default values** for most arguments. 

```{r collapse=TRUE, error=TRUE}
f1 <- function(a, b = 2*a) c(a,b)
f1(3)
f1(3,10)
```

Sometimes reasonable default values might be complicated and lengthy expressions and difficult to add in the function definition. In those cases the **default** is computed within the function only if the argument was not user specified. To check if an argument is missing, the function `missing` is used. 

```{r collapse=TRUE, error=TRUE}
f1 <- function(a, b) 
  {
  if (missing(a)) a <- "default"
  if (missing(b)) b <- "default"
  list(a=a, b=b)
  }

str(f1())
str(f1(1,2))
```

In R it is however more common to set in that case arguments to NULL as default and then check inside the function if the argument is NULL using `is.null`.

```{r collapse=TRUE, error=TRUE}
f2 <- function(a = NULL, b = NULL) 
  {
  if (is.null(a)) a <- "default"
  if (is.null(b)) b <- "default"
  list(a=a, b=b)
  }

str(f2())
str(f2(1,2))
```

### The `...` argument

There is a special argument called **`...`**. This argument **collects** all arguments not matched using the previous rules. These are then usually **passed on** to other functions. The `...` argument is especially used writing so called *wrapper* functions for existing ones where only a few arguments are assigned. E.g. in `plot()` most "tuning" parameters like colors, labels and axes parameters are specified via `...`.

This flexibility comes at a cost:

1. One needs to read the documentation to find out what `...` is passed on to and which arguments are meaningful.
2. Misspelled arguments do not cause errors usually, as they are collected into the `...`.


```{r collapse=TRUE, error=TRUE}
MyLocation <- function(x, loc = "mean", ...) 
  {
  location <- match.arg(loc, c("median", "mean"))
  LABEL <- location
  location <- match.fun(location)
  res <- location(x, ...)
  attr(res, "location") <- LABEL
  res
  }

set.seed(1)
MyData <- rnorm(50)
MyDataNA <- c(MyData, NA)

MyLocation(MyData)

MyLocation(MyData, "med")

MyLocation(MyDataNA, na.rm=TRUE)

MyLocation(MyDataNA, "med", na.rm=TRUE)

MyLocation(MyData, trim=0.2)

MyLocation(MyData, "med", trim=0.2)
```


A problem with `...` occurs when there are several functions to which you would like pass on arguments from `...` - because they must be valid arguments of all of them and get the same value in each function.

It is then often easier to write a function which which as input takes lists with the possible inputs to passed on to the functions in question and use then `do.call`.


### Function returns

Functions in general can return only **one object** as a rule. Which is, however, not a real restriction as all the desired output can be collected into a list. If no return object was explicitely assigned, the last expression evaluated in a function is returned by default. Whenever the function `return(object)` is called within a function, the function is terminated and `object` is returned.


```{r collapse=TRUE, error=TRUE}
f1 <- function(x) 
{
if (x < 0) return("not positive")
if (x < 5) {
   "between 0 and 5"
} else {
 "larger than 5" 
}
  
}
f1(-1)
f1(1)
f1(10)
```


It is possible to return objects from a function call which are not printed by default using the `invisible()` function. Invisible output can be assigned to an object and/or forced to be printed by putting the function call between round parentheses.


```{r collapse=TRUE, error=TRUE}
f1 <- function() 1 
f2 <- function() invisible(1)

f1()
f2()

resf2 <- f2()
resf2
(f2())
```


### Lexical Scoping

This term describes a set of rules that tells R how to find values from names, aka lexical scoping. Lexical scoping looks up symbol values based on how functions were nested when they were created, not how they are nested when they are called. With lexical scoping, you don’t need to know how the function is called to figure out where the value of a variable will be looked up. You just need to look at the function’s definition.

There are four basic principles behind R’s implementation of lexical scoping:

* name masking
* functions vs. variables
* a fresh start
* dynamic lookup

#### Name masking

This principle is best explained by means of examples

```{r collapse = TRUE}
f <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
f()
rm(f)
```

If a name isn’t defined inside a function, R will look one level up.

```{r collapse = TRUE}
x <- 2
g <- function() {
  y <- 1
  c(x, y)
}
g()
rm(x, g)
```

The same rules apply if a function is defined inside another function: look inside the current function, then where that function was defined, and so on, all the way up to the global environment, and then on to other loaded packages. Run the following code in your head, then confirm the output by running the R code.

```{r collapse = TRUE}
x <- 1
h <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z)
  }
  i()
}
h()
rm(x, h)
```

#### Function vs. variables

The same principles apply regardless of the type of associated value — finding functions works exactly the same way as finding variables:

```{r collapse = TRUE}
l <- function(x) x + 1
m <- function() {
  l <- function(x) x * 2
  l(10)
}
m()
```

For functions, there is one small tweak to the rule. If you are using a name in a context where it’s obvious that you want a function (e.g., f(3)), R will ignore objects that are not functions while it is searching. In the following example n takes on a different value depending on whether R is looking for a function or a variable.

```{r collapse = TRUE}
n <- function(x) x / 2
o <- function() {
  n <- 10
  n(n)
}
o()
```

However, using the same name for functions and other objects will make for confusing code, and is generally best avoided.

#### A fresh start

What happens to the values in between invocations of a function? What will happen the first time you run this function? What will happen the second time? (If you haven’t seen `exists()` before: it returns `TRUE` if there’s a variable of that name, otherwise it returns `FALSE`.)

```{r collapse = TRUE}
j <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  a
}
j()
j()
a <- 2
j()
rm(j)

```


You might be surprised that it returns the same value, 1, every time. This is because every time a function is called, a new environment is created to host execution. A function has no way to tell what happened the last time it was run; each invocation is completely independent.



#### Dynamic lookup 

Lexical scoping determines where to look for values, not when to look for them. R looks for values when the function is run, not when it’s created. This means that the output of a function can be different depending on objects outside its environment:

```{r collapse = TRUE}
f <- function() x
x <- 15
f()

x <- 20
f()

```

You generally want to avoid this behaviour because it means the function is no longer self-contained. This is a common error — if you make a spelling mistake in your code, you won’t get an error when you create the function, and you might not even get one when you run the function, depending on what variables are defined in the global environment.

One way to detect this problem is the `findGlobals()` function from `codetools`. This function lists all the external dependencies of a function: 

```{r collapse = TRUE}
f <- function() x + 1
codetools::findGlobals(f)
```


> Excercise D: What does the following code return? What do the different cs stand for?

```{r eval = FALSE}
c <- 10
c(c = c)
```

> Excercise E: What does the following function return?

```{r eval = FALSE}
f <- function(x) {
  f <- function(x) {
    f <- function(x) {
      x ^ 2
    }
    f(x) + 1
  }
  f(x) * 2
}

f(10)
```


## Operators

R contains a number of operators, which can grosso modo be categorized into arithmetic operators and  logical operators. 

### Assignment Operator

One of the most fundamental operators is the assignment operator.

```{r collapse = TRUE}
a <- c(1,2,3)
```

### Arithmetic Operators

The R console can be used as a scientific calculator.

`+`, `-`, `*`, `/`, `^`, `%%`

```{r collapse = TRUE}
1.2 * 5^(-1.2*2)

26 %% 5

```


### Logical Operators

`!`, `&`, `&&`, `|`, `||`, `<`, `>`, `==`, `!=`, `<=`, `>=`


```{r collapse = TRUE}

x <- c(1:10)

# which elements are greater than 8 or less than 5
x[(x>8) | (x<5)]

# every logical operator returns a vector with the same length as x with a TRUE/FALSE
x > 8

!(x > 8)

x < 5

x > 8 | x < 5

x > 8 && x < 5


```


### Operators are Functions!

Really, **every operation in R is a function call**. Even

 `+`, `-`, `*`, `if`, `else`, `while`, `[`, `$`, `{`, ...
 
are functions. In the following note that the backtick `'` lets you refer to functions or variables which have reserved or illegal names. Being allowed to use operators makes the code much more readible for human beings. To demonstrate this, check the following code:
 
 
```{r collapse = TRUE}
'+'(2,3)

'for'(i, 1:2, print("'for' is a function"))

'if'(i==1, print("'if' is a function too"), print("as well as 'else'"))

'<-'(x, 1:5)
'['(x,3)

```

Knowing that these operators are actually functions is especially useful when you want to use them in a looping structure. 

```{r collapse=TRUE, error=TRUE, asis=TRUE}
L1 <- list(a=1:3, b=70:77)
L1

sapply(L1, '[', 2)

sapply(L1, '+', 10)

```



