---
title: "Excercise Sheet"
output: pdf_document
---

1. Open RStudio and create a new RMarkdown File (.Rmd) with the follwing YAML parameter:
  + Title: The Iris Data Set
  + Document Type
  + Default Output Format: PDF

2. Create an R chunk called: load

``` {r load, eval = FALSE, echo = TRUE}
data("iris")
library(c("stargazer", "xtable"))

```

3. Make horizontal rule under the title

4. Create the subtitle: Description

5. Print a summary of the dataset using
``` {r summary1, eval = FALSE, echo = TRUE}
summary(iris)

```

6. Create a nicer summary using 
``` {r summary2, eval = FALSE, echo = TRUE}
options(xtable.comment = FALSE)
xtable::xtable(summary(iris), type = "latex", comment = FALSE, caption = "Summary of the data set")
```

7. Create a new section called Scatterplot Matrix and plot
``` {r plot, eval = FALSE, echo = TRUE}
plot(iris)
```

8. Perform a logistic regression analysis and produce a neat table with the output
``` {r logreg, eval = FALSE, echo = TRUE}
iris[['Is.Versicolor']] <- as.numeric(iris[['Species']] == 'versicolor')
iris[['Is.Virginica']] <- as.numeric(iris[['Species']] == 'virginica')
fit.1 <- glm(Is.Versicolor ~ Petal.Length + Sepal.Length, data = iris)
fit.2 <- glm(Is.Virginica ~ Petal.Length + Sepal.Length, data = iris)
#summary(fit)
output <- capture.output(stargazer(fit.1, fit.2, title = 'Regression Results', 	summary=FALSE, header=FALSE))
cat(output)
```
	
9. Include the references using bibliography.bib

