---
title: "Pipe Operator"
author: "Fabian Schroeder"
date: "April 11, 2018"
output: html_document
---

## The Pipe Operator `%<%`

Pipes are a powerful tool for expressing a sequence of multiple operations in a way that is easily comprehensible by the human mind. Its introduction is one of the biggest changes to the R language in recent years. 

The pipe, %>%, was introduced to the R language by the magrittr package by Stefan Milton Bache. It existed in other programming languages before and is, thus, not an R specific method. However, it is in particulary useful for data science. Even though it was not originally developed for the `tidyverse` it is now an essential part of it and the `magrittr` package is automatically loaded by loading the `tidyverse`.

``` {r}
library(magrittr)
```

Let us start with an introductory example by the author himself in the following blog:

[blog](https://www.r-statistics.com/2014/08/simpler-r-coding-with-pipes-the-present-and-future-of-the-magrittr-package/)

```{r collpse = TRUE, fig.align="center"}
library(babynames) # data package
library(dplyr)     # provides data manipulating functions.
library(magrittr)  # ceci n'est pas un pipe
library(ggplot2)   # for graphics

babynames %>%
    filter(name %>% substr(1, 3) %>% equals("Ste")) %>%
    group_by(year, sex) %>%
    summarize(total = sum(n)) %>%
    qplot(year, total, color = sex, data = ., geom = "line") %>%
    add(ggtitle('Names starting with "Ste"')) %>%
    print

```

even without knowing much about magrittr (or even R) reading this chunk of code is pretty easy — like a recipe, and not a single temporary variable is needed. It’s almost like

1. take the baby data, then
2. filter it such that the name sub-string from character 1 to 3 equals "Ste", then
3. group it by year and sex, then
4. summarize it by computing total sum for each group, then
5. plot the resuls, coloring by sex, then
6. add a title, then
7. print it to the canvas.

The order in which you’d think of these steps is the same as the order in which they are written, and as the order in which they are executed.
 The alternative would be to use either a bunch of variables, or to have a nasty string of nested functions calls starting with print at the very left, babynames somewhere in the middle, and the remaining arguments and values scattered around.


The point of the pipe is to help you write code in a way that easier to read and understand. To see why the pipe is so useful, we’re going to explore four ways of writing the same code.

* Save each intermediate step as a new object.
* Overwrite the original object many times.
* Compose functions.
* Use the pipe.

We’ll work through each approach, showing you the code and talking about the advantages and disadvantages. 

### Save each intermediate step

The simplest approach is to save each step as a new object:

``` {r eval = FALSE}
data_1 <- func_1(data)
data_2 <- func_2(data_1)
data_3 <- func_3(data_2)

```

The main downside of this form is that it forces you to name each intermediate element. If there are natural names, this is a good idea, and you should do it. But many times, like this in this example, there aren’t natural names, and you add numeric suffixes to make the names unique. That leads to two problems:

* The code is cluttered with unimportant names
* You have to carefully increment the suffix on each line.

You may also worry that this form creates many copies of your data and takes up a lot of memory. Surprisingly, that’s not the case. First, note that proactively worrying about memory is not a useful way to spend your time: worry about it when it becomes a problem (i.e. you run out of memory), not before. Second, R isn’t stupid, and it will share columns across data frames, where possible. 

### Overwrite the original object

Instead of creating intermediate objects at each step, we could overwrite the original object:

``` {r eval = FALSE}
data <- func_1(data)
data <- func_2(data)
data <- func_3(data)

```

This is less typing (and less thinking), so you’re less likely to make mistakes. However, there are two problems:

* Debugging is painful: if you make a mistake you’ll need to re-run the complete pipeline from the beginning.
* The repetition of the object being transformed (we’ve written foo_foo six times!) obscures what’s changing on each line.

### Compose functions

Another approach is to abandon assignment and just string the function calls together:

```{r eval=FALSE}
func_3(
  func_2(
    func_3(data)
  )
)
```

Here the disadvantage is that you have to read from inside-out, from right-to-left, and that the arguments end up spread far apart (evocatively called the dagwood sandwhich problem). In short, this code is hard for a human to consume.

### Use the pipe

```{r eval=FALSE}
data %>%
  func_1(arg_1) %>%
  func_2(arg_2) %>%
  func_3(arg_3)
```

### Review the example

Let us have a closer look at the example

```{r eval=FALSE}
babynames %>%
    filter(name %>% substr(1, 3) %>% equals("Ste")) %>%
    group_by(year, sex) %>%
    summarize(total = sum(n)) %>%
    qplot(year, total, color = sex, data = ., geom = "line") %>%
    add(ggtitle('Names starting with "Ste"')) %>%
    print

```


The example illustrates a few feature of %>%. Firstly, the dplyr functions filter, group_by, and summarize all take as first argument a data object, and as default this is where %>% will place its left-hand side. The babynames data is thus inserted as first argument in the call to filter. When the filtering is done, the result is passed as the first argument to group_by, and similarly for summarize. However, one is not always so fortunate that a function is designed to accept the data (or whatever you might be piping along) as its first argument (the dplyr functions are designed with %>% operations in mind). This is the case with e.g. qplot, but note the data = . argument. This tells %>% to place the left-hand side there, and not as the first argument. This is a simple and natural way to accommodate the lack of consistency of function signatures, and allows the left-hand side to go anywhere in the call on the right-hand side. You may also have noted that print is used without parentheses; this is to make the code even cleaner when only one the left-hand side is needed as input. Finally, note that %>% can be used in a nested fashion (a separate chain is found within the filter call) and that magrittr has aliases for commonly used operators, such as add for `+` and equals for `==` used above. These make pipe chains more readable (not necessarily shorter).
 


