---
title: "library_vs_require"
author: "Fabian Schroeder"
date: "April 18, 2018"
output: html_document
---

## Difference between `library()` and `require()`

`library(package)` and `require(package)` both load the package with name package and put it on the search list. `require` is designed for use inside other functions; it returns `FALSE` and gives a warning (rather than an error as `library()` does by default) if the package does not exist. Both functions check and update the list of currently loaded packages and do not reload a package which is already loaded. (If you want to reload such a package, call detach(unload = TRUE) or unloadNamespace first.) If you want to load a package without putting it on the search list, use requireNamespace.


```{r }
test <- library("abc")
test

test <- require("abc")
test
```

You can use it in:

```{r }
if(require("lme4")){
    print("lme4 is loaded correctly")
} else {
    print("trying to install lme4")
    install.packages("lme4")
    if(require(lme4)){
        print("lme4 installed and loaded")
    } else {
        stop("could not install lme4")
    }
}

```




