---
title: "107.106 AKSTA Statistical Computing"
subtitle: "Graphics with R" 
author: "Fabian Schroeder"
date: "SS 2018"
output: html_document
---

 
 
 




## Plotting example I
The following plot should give an impression of the colours, symbols and point sizes in R.

```{r}
plot(1,1,xlim=c(1,10),ylim=c(0,5),type="n")
points(1:9,rep(4.5,9),cex=1:9,col=1:9,pch=0:8)
text(1:9,rep(3.5,9),labels=paste(0:9),cex=1:9,col=1:9)
points(1:9,rep(2,9),pch=9:17)
text((1:9)+0.25,rep(2,9),paste(9:17))
points(1:8,rep(1,8),pch=18:25)
text((1:8)+0.25,rep(1,8),paste(18:25))
```
 
## Plotting example II

This plot is about putting two figures into one window.
```{r}
set.seed(1) # to make it reproducible
x<-rnorm(80)
breaks<-(-ceiling(max(abs(x))):ceiling(max(abs(x))))
ylim=c(0,0.6)
par(mfrow=c(1,2))
hist(x,freq=F,breaks=breaks,add=F,ylim=ylim)
curve(dnorm(x),add=TRUE)
plot(density(x),type="l",main="Kernel Density of x"
     ,ylim=ylim)
hist(x,freq=FALSE,add=TRUE)
par(mfrow=c(1,1))
```
 

## Plotting example III

This example shows how to insert a legend interactively into a scatter plot which has regression lines added. And also how to deal with long titles.

```{r, eval=FALSE}
library(MASS)
data(whiteside)
attach(whiteside)
After<-subset(whiteside,Insul=="After")
Before<-subset(whiteside,Insul=="Before")
lm.after<-lm(Gas~Temp,data=After)
lm.before<-lm(Gas~Temp,data=Before)
plot(Temp,Gas,pch=as.numeric(Insul),
main="Gas comsumption of Derek Whiteside's\n house
before and after insulation")
abline(lm.before)
abline(lm.after)
legend(locator(1),legend=c("Before insulation",
"After insulation"),pch=c(1,2))
```


## Plotting example IV

This example show parallel boxplots which are made using a model formulae. The ticks on the right and left side of the plot indicate the observations.   

The data used are the same as in the example before.

```{r, eval=FALSE}
boxplot(Gas~Insul,col=c("grey50","grey80"))
rug(Gas[Insul=="Before"],side=2)
rug(Gas[Insul=="After"],side=4)
```
 

## Plotting example IV - the plot

```{r, echo=FALSE}
library(MASS)
data(whiteside)
attach(whiteside)
After<-subset(whiteside,Insul=="After")
Before<-subset(whiteside,Insul=="Before")
lm.after<-lm(Gas~Temp,data=After)
lm.before<-lm(Gas~Temp,data=Before)
boxplot(Gas~Insul,col=c("grey50","grey80"))
rug(Gas[Insul=="Before"],side=2)
rug(Gas[Insul=="After"],side=4)
```
 
